%% ============================================================
%% Chapter 1 — Introduction
%% ============================================================

\chapter{Introduction}\label{chapter:1_Introduction}

\section{Context}
\label{sec:intro:context}

Mandarin Chinese has approximately 1.18~billion native and second-language speakers worldwide, of whom roughly 194~million speak it as a second language~\cite{ethnologue2025}. Unlike most Indo-European languages, Mandarin is a tonal language: the pitch contour of each syllable is phonemically distinctive, so a change in tone changes the word itself~\cite{jongman2006tones}. Standard Mandarin employs four lexical tones: a high-level tone~(T1), a rising tone~(T2), a dipping tone~(T3), and a falling tone~(T4), plus a reduced neutral tone that occurs in specific unstressed syllables~\cite{zahner2022toneexperience}. A canonical illustration is the syllable \textit{ma}, which means ``mother'' with a high-level tone (T1), ``hemp'' with a rising tone (T2), ``horse'' with a dipping tone (T3), and ``scold'' with a falling tone (T4)~\cite{jongman2006tones}. None of the target words used in this thesis takes the neutral tone as its canonical realisation, so the neutral tone does not occur in our data. The additional label introduced in Chapter~\ref{chapter:3_Proposal} is reserved for a separate phenomenon: learner productions that the expert hears as outside any of the four canonical tones.

The tonal nature of Mandarin poses a substantial challenge for second-language~(L2) learners whose first language is non-tonal. In non-tonal languages such as Polish, English, or French, pitch variation serves only a prosodic function, signalling questions, emphasis, or discourse boundaries, rather than a lexical one. Consequently, learners from these linguistic backgrounds tend to underperceive tonal contrasts and produce pitch contours that deviate systematically from native targets~\cite{hao2012}. Research on L2 tone acquisition has identified certain tone pairs as particularly problematic: the T2/T3 contrast, where both tones begin in the low-pitch region, is consistently the most confusable for non-tonal-L1 learners~\cite{hao2012}. A recent study of Polish learners of Mandarin reported systematic differences in both pitch height and contour shape relative to native speakers across disyllabic tonal combinations~\cite{chu2024statistical}. Polish learners are therefore a natural target population for evaluating automatic Mandarin tone-training methods.

Computer-Assisted Pronunciation Training~(CAPT) systems aim to address this need by providing automated, scalable feedback on learner pronunciation~\cite{neri2002,rogerson2021}. A CAPT system typically comprises an automatic speech recognition or classification component that analyses the learner's production, coupled with a feedback mechanism that communicates results. In Mandarin tone assessment specifically, early classifiers adopted convolutional neural network~(CNN) architectures over hand-designed spectral inputs, such as MFCCs combined with autoencoder-pretrained feature kernels~\cite{chen2016cnn} or fixed-size mel-spectrogram images~\cite{gao2019tonenet}, achieving high accuracy on isolated syllables under controlled conditions. These models demonstrate strong performance on clean, single-syllable input, but their reliance on hand-designed acoustic features and from-scratch training leaves open how well they transfer to learner speech across recording conditions, proficiency levels, and population-specific error patterns.

Self-supervised learning~(SSL) offers a different methodological route. Models such as Wav2Vec~2.0~\cite{baevski2020} and HuBERT~\cite{hsu2021} are pre-trained on large amounts of unlabelled speech and then adapted to downstream tasks with comparatively little labelled data. For Mandarin specifically, layer-wise probing has shown that intermediate SSL representations encode information relevant to lexical tone and other suprasegmental categories, and that task-specific fine-tuning further sharpens this signal~\cite{yuan2023,delafuente2024}. SSL-based classifiers thus offer a natural alternative to earlier CNN approaches, particularly when learner speech is acoustically variable and labelled data are scarce.

\section{Identification of Gaps}
\label{sec:intro:gaps}

The literature reviewed above leaves three issues unresolved for the present study.

\textbf{Population gap.} Most tone-classification systems reviewed in Chapter~\ref{chapter:2_StateOfTheArt} were developed and evaluated on native Mandarin speech~\cite{chen2016cnn,gao2019tonenet}. Publicly available learner-oriented resources remain limited, and the non-native corpus most directly relevant to pronunciation assessment is OMPAL, which records French~L1 rather than Polish learners~\cite{hsieh2025ompal}. For Polish learners of Mandarin, the available studies are acoustic and pedagogical in focus: Chu et~al.~\cite{chu2024statistical} provide a quantitative analysis of disyllabic tone production, and Zajdler~\cite{zajdler2021} examines tonal production in read speech, but neither study develops or evaluates an automatic tone-classification model. To the best of the author's knowledge, no published study has evaluated modern Mandarin tone classifiers on speech produced by Polish learners.

\textbf{Model comparison gap.} While both CNN-based and SSL-based approaches have been applied to Mandarin tone classification independently, no systematic comparison of these paradigms has been conducted on non-native learner speech. Furthermore, SSL models can be pre-trained on different data, including native Mandarin corpora, mixed native-and-learner corpora, or multilingual checkpoints, and the effect of these pre-training strategies on downstream tone classification accuracy for L2 speakers remains unexplored.

\textbf{Expert alignment gap.} Mandarin tone classifiers reported in the literature are usually evaluated against \emph{intended-tone} labels, that is, the tone prescribed by the reading prompt or dictionary entry. For native-speaker corpora this is a reasonable proxy, because the intended and realised tones almost always coincide. For learner speech, however, the two may diverge. In the literature reviewed for this thesis, I did not find a Mandarin tone-classification study that evaluates learner recordings against per-recording expert perceptual tone labels of this kind.

\section{Goals and Scope}
\label{sec:intro:goals}

The principal goal of this thesis is to develop an automatic pronunciation assessment model for Mandarin Chinese learner speech that scores production along two largely independent dimensions: \emph{phonetic accuracy}, the correctness of the consonant--vowel realisation considered apart from tone, and \emph{tonal accuracy}, the lexical tone actually produced. Because these dimensions reflect distinct phonological properties, and were annotated as such by a native expert, the assessment is framed as two independent classification tasks trained on a shared learner corpus. The implemented experiments target the tonal-accuracy model, and the thesis specifies how it can be combined with the corresponding phonetic-accuracy component in a CAPT pipeline.

This overarching goal is pursued through the following specific tasks:

\begin{enumerate}
  \item \textbf{Phonetic accuracy model.} Define the segmental assessment task in which a model receives a learner recording of a known target word and predicts whether the consonant--vowel realisation is acceptable independently of tone. An implementation of this component on the same Polish learner corpus has been carried out as a parallel project~\cite{brzeski2025mprp}; this thesis does not re-implement the segmental classifier but treats it as the partner component to the tonal-accuracy model developed here.
  \item \textbf{Tonal accuracy model.} Develop and evaluate a model that classifies the realised tone of a learner recording as \(T_1\), \(T_2\), \(T_3\), \(T_4\), or \(T_{\mathrm{NS}}\), where \(T_{\mathrm{NS}}\) denotes a non-standard production outside the four canonical lexical tones.
  \item \textbf{Comparative evaluation of CNN-based and SSL-based architectures.} Compare a ToneNet-style CNN baseline with Wav2Vec~2.0 and HuBERT systems on the expert-labelled learner corpus.
  \item \textbf{Effect of pre-training and adaptation strategy.} Examine whether learner-only fine-tuning, native Mandarin AISHELL-1 adaptation~\cite{bu2017aishell}, or Mandarin-specific SSL checkpoints provide the strongest basis for learner tone assessment.
  \item \textbf{Alignment with expert judgement.} Evaluate model outputs against per-recording expert perceptual labels rather than only against prompt-derived intended tones.
  \item \textbf{Integration guidance for CAPT.} Identify how the tonal classifier can be used together with a phonetic-accuracy classifier in a future pronunciation-feedback system.
\end{enumerate}

\section{Outline of the Thesis}
\label{sec:intro:outline}

The thesis is organised as follows:

\begin{itemize}
  \item Chapter~\ref{chapter:2_StateOfTheArt} reviews the relevant literature on Mandarin tone phonology, second-language tone acquisition, computer-assisted pronunciation training, and self-supervised speech modelling.
  \item Chapter~\ref{chapter:3_Proposal} defines the research problem, states the main assumptions and requirements, and specifies the acceptance criteria for the study.
  \item Chapter~\ref{chapter:4_Design} describes the system architecture, model configurations, and data-processing pipeline used in the implementation.
  \item Chapter~\ref{chapter:5_Experiments} presents the experimental setups, evaluation procedure, quantitative results, and comparative analysis.
  \item Chapter~\ref{chapter:6_Summary} summarises the main findings, discusses limitations, and outlines directions for future work.
\end{itemize}
