%% ============================================================
%% Chapter 6 — Summary and Future Work
%% ============================================================

\chapter{Summary and Future Work}\label{chapter:6_Summary}

\section{Summary}
\label{sec:summary:summary}

This thesis studied automatic Mandarin tone classification for Polish learners
using both a ToneNet-style convolutional baseline and fine-tuned self-supervised
learning~(SSL) speech models. The task was framed as five-way tone classification over
\(T_1\), \(T_2\), \(T_3\), \(T_4\), and \(T_{\mathrm{NS}}\), where
\(T_{\mathrm{NS}}\) denotes a non-standard learner production outside the four
canonical Mandarin tones. The target label was the expert annotation of the
learner's produced tone, rather than the intended prompt tone. This makes the
evaluation closer to computer-assisted pronunciation training, where the
relevant question is whether a system agrees with a human assessment of the
learner production.

The implemented pipeline converted the original learner assessment material
into two experimental representations. For the SSL systems, MFA
alignments were used to build frame masks over 11\,323
usable utterances from 678 speakers. The train, development, and test partitions
were speaker-disjoint, with 8\,577, 464, and 2\,282 utterances respectively. For
the CNN baseline, low-frequency mel-spectrogram images were generated and split
into 9\,546 training, 1\,194 validation, and 1\,194 test images. The CNN result
therefore serves as a ToneNet-style reference, while the SSL systems
provide the main speaker-disjoint comparison.

Six complete model configurations were evaluated. The CNN baseline classified
mel-spectrogram images. The Wav2Vec~2.0 experiments compared direct learner
fine-tuning, AISHELL-1 intermediate tone training followed by learner
fine-tuning, and a Mandarin-pretrained Wav2Vec~2.0 checkpoint. The HuBERT
experiments compared a Mandarin-pretrained frame-level model with majority
voting against an utterance-level attention model trained with focal loss. The
strongest overall system was Chinese HuBERT-Large with frame-level
cross-entropy, which reached 80.28\% accuracy, 73.37\% macro-F1, and 80.00\%
weighted F1 on the learner test set. Among the Wav2Vec~2.0 systems, the
strongest macro-F1 came from the AISHELL-1 label1 route at 71.46\%, with the
three AISHELL-1 intermediate variants converging to 71.41--71.46\% macro-F1.

\section{Fulfilment of the Project Goals}
\label{sec:summary:goals}

Within the implemented tonal-assessment scope, the implementation satisfies the
main technical goals defined for the project.
First, the thesis produced a complete learner-speech preparation workflow that
links expert labels, audio files, MFA speech masks, and speaker identities.
Second, it reproduced a CNN baseline for comparison with prior
image-based tone-classification work. Third, it implemented multiple SSL
fine-tuning routes using Wav2Vec~2.0 and HuBERT, including native Mandarin
intermediate training and Mandarin-pretrained checkpoints. Fourth, it evaluated
the SSL models under a speaker-disjoint protocol and reported both aggregate and
class-level metrics, which is necessary for an imbalanced learner corpus.
Finally, it evaluated the models against expert perceptual labels and described
how the tonal classifier could later be combined with a parallel
phonetic-accuracy classifier in a CAPT pipeline.

The acceptance criteria are therefore met at the research-prototype level. The
system is not yet a deployable pronunciation-training application, and the
reported labels come from the available expert annotation rather than a new
multi-rater study. These points limit the scope of the conclusions, but they do
not undermine the central comparison: modern SSL speech models,
especially Chinese HuBERT-Large, provide a stronger foundation than the
spectrogram CNN baseline for Mandarin tone classification in this Polish learner
corpus.

\section{Limitations}
\label{sec:summary:limitations}

The first limitation is corpus size and balance. The final SSL test set is large
enough to compare model families, but the distribution is uneven across tones.
\(T_3\) and \(T_{\mathrm{NS}}\) have fewer and more variable examples than the
dominant classes, which makes their F1 scores sensitive to relatively small
changes in the confusion matrix.

The second limitation is supervision granularity. The learner labels are
utterance-level expert ratings copied to the MFA-selected speech frames. This is
a practical way to train frame-level SSL models, but it does not tell the model
where inside the syllable the decisive tonal evidence occurs. Native AISHELL-1
frame labels are more detailed, but they are still transferred from read native
speech to non-native learner productions.

The third limitation is evaluation comparability for the CNN. The CNN baseline
uses a random image split, following the ToneNet-style workflow implemented in
the notebooks. The SSL experiments use speaker-disjoint splits. The CNN result
is useful as a baseline reference, but a future fully controlled comparison
should train the image model under the same speaker-disjoint split.

The fourth limitation is that the thesis evaluates automatic label agreement,
not pedagogical effectiveness. A model that predicts expert tone labels can be a
component of pronunciation feedback, but a deployed learning tool would also
need calibrated confidence, interpretable feedback, latency constraints, and
validation with learners and instructors.

\section{Future Work}
\label{sec:summary:future}

Future work should focus first on improving the accuracy and robustness of the
best SSL tone classifier. The strongest model still makes many errors on \(T_3\)
and \(T_{\mathrm{NS}}\), so additional experiments should target these classes
directly through better class balancing, class-aware losses~\cite{lin2017focal},
calibrated decision thresholds, and stronger data augmentation. Model selection
should also continue to use macro-F1 or another class-balanced criterion rather
than accuracy alone, because the practical CAPT value of the system depends on
its behaviour on difficult and less frequent learner productions.

A second direction is to make the tone classifier prompt-aware. The present
systems predict only from the learner waveform and reduce frame-level decisions
to a single utterance label by majority voting. A feedback system, however,
knows the word or phrase that the learner was asked to pronounce and the
expected lexical tone of the target syllable. Future models should therefore
condition prediction on this context, for example by providing the target word,
the expected tone number, or a syllable-level prompt representation together
with the acoustic frames. This would allow the system to judge whether the
frames that should carry the target tone actually support the expected contour,
rather than treating all speech frames as equally informative votes.

The frame-level decision rule should also be redesigned for feedback. Instead
of assigning the utterance label by majority vote over all masked frames, future
work could use tone-bearing phone intervals, syllable-level alignment,
CTC-style objectives~\cite{fan2023ctcbert}, attention over the expected target
syllable, or auxiliary pitch-contour targets. These methods would make it
possible to identify the part of the recording that caused an incorrect tone
decision and to provide more precise feedback than a single global label.

Finally, the improved tone model should be evaluated inside a prototype CAPT
pipeline together with the parallel phonetic-accuracy
classifier~\cite{brzeski2025mprp}. Such a prototype should combine the predicted
tone, the expected prompt tone, model confidence, and the relevant frame or
syllable evidence before deciding whether to mark a production as correct,
incorrect, or uncertain. This would move the thesis model from expert-label
classification toward learner-facing feedback while keeping the feedback tied to
the acoustic evidence used by the classifier~\cite{rogerson2021,neri2002}.
