%% ============================================================
%% Abstract (English)
%% ============================================================
\addcontentsline{toc}{chapter}{Abstract}
\chapter*{Abstract}

Mandarin Chinese is a tonal language in which syllable-level pitch patterns distinguish lexical meaning, making tone production a central challenge for learners whose first language is non-tonal. This thesis investigates automatic Mandarin tone classification for Polish L1 learners as the assessment component of a future computer-assisted pronunciation training system. Pronunciation quality is decomposed into two largely independent dimensions, phonetic accuracy of the consonant--vowel realisation and tonal accuracy of the produced tone; the phonetic component is addressed in a parallel project on the same learner corpus, while this thesis develops and evaluates the tonal-accuracy classifier. Six modelling setups are evaluated: a ToneNet-style convolutional neural network (CNN) baseline using mel-spectrogram images, three Wav2Vec 2.0 variants with different adaptation routes, and two Chinese HuBERT-Large variants with different prediction heads.

The experiments use an expert-labelled learner corpus of Mandarin recordings produced by Polish speakers. The SSL systems are evaluated on a speaker-disjoint split containing 8,577 training utterances, 464 development utterances, and 2,282 test utterances; the CNN is retained as a ToneNet-style reference under an independent speaker-disjoint image split containing 9,040 training images, 487 validation images, and 2,407 test images. Across the evaluated setups, Chinese HuBERT-Large with frame-level cross-entropy achieves the highest test accuracy, macro-F1, and weighted F1, reaching 80.28\% accuracy, 73.37\% macro-F1, and 80.00\% weighted F1. Among the Wav2Vec 2.0 routes, the strongest macro-F1 is obtained by the AISHELL-1 label1 intermediate-adaptation route, reaching 71.46\%, while the three AISHELL-1 variants cluster between 71.41\% and 71.46\%. These results suggest that frame-level supervision over SSL representations is effective for learner-tone classification and that non-canonical or contour-variable productions remain the main source of residual errors.

\section*{Keywords:}
\noindent Mandarin tone classification, computer-assisted pronunciation training, self-supervised learning, Wav2Vec 2.0, HuBERT, Polish L1 learners
